Fine-scale Twitter Big Data Reveals Neighborhood Inequalities in Mental Health Effects of High Temperatures

### 2021-05-19 doc/temp-sentiment.tex
Updates from Overleaf

\begin{abstract}
%Limit 150 words: https://www.nature.com/nclimate/about/content
%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health.  Indeed, traditional analyses of heat and mental health outcomes using data collected by public agencies have found strong associations between elevated temperatures and outcomes such as suicides and mental-health related hospitalizations.  However, these studies, which typically use data reported at the city or county level, have found no difference in vulnerability based on income or race.  We overturn this finding and show that there are stark differences in vulnerability using a novel data source: expressed sentiment in a quarter of a billion geolocated tweets, matched with prevailing weather as well as neighborhood economic and demographic conditions.  We find that increased temperatures worsen expressed sentiment in all areas, but that this effects is much stronger in poor and black neighborhoods.
\end{abstract}
\begin{document}

@@

\section*{Main}
%Overview
Climate change is impacting many aspects of human well-being and a large body of research has assessed the ongoing impacts on climate change on outcomes like agriculture, economic growth, and physical health \cite{pachauri2014climate}.  However, academic and medical researchers have highlighted the relative paucity of empirical studies on the potential impacts of climate change on mental health \cite{Berry2018Apr, hayes_climate_2018}.  While mental health is under-studied, it makes up 13\% of the global burden of disease, representing a major share of total human suffering \cite{Collins2011Jul}.  Researchers have begun to answer the call more more research, finding associations between climate disasters and post-traumatic stress \cite{Waite2017Dec, Raker2019Dec}, deteriorating environmental conditions and a sense of "ecological grief" \cite{Cunsolo2018Apr}, as well as higher temperatures and a variety of mental health impacts \cite{baylis_weather_2018, Mullins2019Dec, Li2020Mar, Obradovich2018Oct}.  Nevertheless, as global warming may reach 1.5\textdegree C within a decade \cite{allen2019technical}, more research is needed into the impacts of heat on mental health, especially in identifying the most vulnerable communities.

%Theoritical: frameworks and pathways
A number of frameworks have been proposed highlighting the pathways by which temperature can affect mental health \cite{Berry2018Apr, Palinkas2020Apr, BerryETAL2010}.  Work on biological mechanisms emphasizes that there are mental health effects from dehydration and other consequences of maintaining stable body temperature in high heat \cite{Lohmus2018Jul, sadiq_impact_2019}.  Additionally, recent studies have found that nighttime temperatures affect sleep quality, with consequences for mental health \cite{Obradovich2017May, Mullins2019Dec}.  It has also been noted that injuries and physical health impacts from heat can have consequences for mental health \cite{Berry2007, WHO2007}.  Finally, higher temperatures affect productivity and income \cite{kjellstrom_impact_2016, Burke2015Nov}, leading to second-order impacts on mental health \cite{Katz1997, CohnETAL2004, BouchamaETAL2007}.

%Empirical: Heat and mental health
Building on these frameworks, an emerging literature has shown that higher temperatures are associated with many indicators of worsened mental health.  Multiple studies have found strong evidence that higher temperatures are associated with increases in suicides in the United States \cite{Burke2018Aug, Mullins2019Dec, Dixon2007May} with other work demonstrating similar relationships in locations around the world \cite{Qi2014Dec, Page2007Aug, Likhvar2011Jan}.  Higher temperatures have also been correlated with increases in hospitalization events related to both general mental health issues \cite{Obradovich2018Oct, Mullins2019Dec} specific conditions like bipolar disorder and schizophrenia \cite{Lee2007Jan, Sung2013Feb} as well as increased mortality among people with mental health disorders \cite{Hansen2008Oct}.

While the effects of heat on mental health are now well established, research conducted at national scales has found surprisingly little heterogeneity in impacts, with consequences for our understanding of vulnerability and adaptation.  For example, focusing on suicides, Burke et al. found ``no significant difference in suicide response to temperature between rich and poor municipalities or counties" \cite{Burke2018Aug}.  Similarly, Mullins et al. find no effect of income as modifier of the effect of temperature on suicides, emergency department visits, or self-reported mental health status \cite{Mullins2019Dec}.  However, one major challenge for these studies is that they rely on data from public health agencies that is aggregated to the municipality or county level.  As a consequence, they are restricted to between-county metrics of vulnerability and cannot take neighborhood effects into account.

These findings of uniform vulnerability to heat from from national studies are surprising, because the poor and disadvantaged are much more exposed to undesirable temperatures.  They are more likely to work outside rather than indoors \cite{Gubernot2014Oct}, more likely to rely on public transportation, bicycle, or walk instead of commute in their own air-conditioned car \cite{Karner2015Dec}, and more likely to live in housing that is less insulated with poorer quality air-conditioning \cite{Samuelson2020Jun}.  Thus, there are strong reasons to expect heterogeneities in the impact of heat on mental health outcomes, and these heterogeneities in vulnerability might be most pronounced within cities rather than between cities.  Indeed, neighborhoods have a large effect on vulnerability for a variety of impacts of heat on physical health  \cite{Belanger2015Mar, Uejio2011Mar}, as well as for the mental health impacts of other types of climate shocks, such as hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}.

While public health data on mental health outcomes is not available at the neighborhood scale, data from the social media platform Twitter can provide an important indicator of mental health at extremely precise spatial and temporal scales.  Many studies have used the mood expressed in tweets as an indicator of mental health \cite{Edo-Osagie2020Jul, Sinnenberg2016Dec}.  Such studies have used Twitter data to identify the onset of Post Traumatic Stress Disorder in individuals, even before their formal diagnosis \cite{Reece2017Oct}, while similar methods with Facebook data have been used to predict the onset of depression \cite{Eichstaedt2018Oct}.  Additionally, in London, day-to-day changes in mental health indicators derived from tweets have been associated with changes in mental health crisis episodes \cite{Kolliakou2020Feb}, while across the US, more positive tweets are associated with a variety of metrics of human well-being \cite{Mitchell2013May}.  Finally, the mood expressed in tweets has been strongly associated with local weather \cite{baylis_weather_2018, hannak_tweetin_2012}, and has been used as evidence to support the temperature-suicide relationship \cite{Burke2018Aug}.  Thus, Twitter provides an indicator of mental health at fine enough spatial scales to examine neighborhood effects on vulnerability to higher temperatures.  We therefore draw on a data set of 243 million geolocated tweets from across the continental United States and covering the period 2009-2019, each paired local weather and neighborhood conditions, to explore how neighborhood conditions affect the relationship between heat and mental heath.

@@

\subsection*{Overall Effect}
We examine the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), an temperature metric that accounts for dry-bulb temperature, humidity, wind speed, and solar radiation to more accurately describe the effects of heat on the human body \cite{budd2008wet}.  Controlling for a variety of fixed effects, we find that, across all tweets, higher temperatures are associated with lower sentiment (See Fig. \ref{fig:wbgt}).  Sentiment is highest at a WBGT of 5\textdegree C (a dry bulb temperature typically around 12\textdegree C/54\textdegree F), and the greatest declines in sentiment are between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F).  Sentiment also declines with colder temperatures, but only slightly.  We run models using multiple sentiment indicators, and find similar results no matter which sentiment metric was used (See Supplement).

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{../res/wbgt.png}
  \caption{Relationship between Wet Bulb Globe Temperature (WBGT) and sentiment in tweets.  As temperatures increase, sentiment declines.}
  \label{fig:wbgt}
\end{figure}

We then explore how the racial and income characteristics of neighborhoods moderate the relationship between WBGT and sentiment (See Fig \ref{fig:hetero}).  We find that neighborhood income strongly moderates the relationship between temperature and sentiment.  As temperatures increase to a moderate heat of 20\textdegree C WBGT (29\textdegree C/84\textdegree F), sentiment decreases for most neighborhoods, but increases in neighborhoods in the 95th income percentile.  These rich areas do not see decreases in sentiment until temperatures are above 20\textdegree C WBGT (29\textdegree C/84\textdegree F), at which point sentiment decreases evenly across income ranges, still leaving a large gap in expressed sentiment from low to high-income neighborhoods.

We also examine patterns of temperature and sentiment for neighborhoods that had a majority population of four broad racial categories - non-Hispanic black, non-Hispanic white, Hispanic of any race, and other, which includes Native American, multi-racial, and Asian-American and Pacific Islander.  While people in neighborhoods of all racial compositions are affected by heat, people in majority black neighborhoods are much more affected than others  (See Fig \ref{fig:hetero}).  Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the sentiment of tweets in majority black neighborhoods decrease four times as much as people in other neighborhoods.  Additionally, at levels of moderate heat from 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority Hispanic neighborhoods have slightly lower sentiment than people in majority white or other neighborhoods, although this gap closes with high temperatures.

Finally, we model the combined effects of race and income in moderating the impacts of heat on mental health, and we find separate effects for these two variables (See Supplement).  In other words, neither race nor income alone account for all the heterogeneity in vulnerability, and neighborhoods that are poor and black are more affected by heat than than neighborhoods that are just poor or just black.

@@

\subsection*{Comparison With Other Events}

Expressed sentiment is a widely used metric to assess mental health and well-being on social media.  While a variety of related metrics can be used to quantify sentiment, we use the VADER metric that was specifically designed for microblogs like Twitter \cite{hutto2014vader}.  To give context to this novel metric, we compare the impacts of heat waves, defined as a change from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), to other events that have large impacts on sentiment (See Fig. \ref{fig:compare}).  First, we examine the change in average sentiment from Saturday to Monday, the highest and lowest days for Twitter sentiment as well as other mental health metrics, such as suicide rates \cite{CDC2021}.  We also examine the decrease in sentiment associated with the two most expensive hurricanes in the last decade in the United States.  For the hurricanes, we compare the mean sentiment of counties affected by Hurricanes Harvey and Sandy on the day of landfall to the mean sentiment exactly one week before.  These hurricanes are associated with mental health effects including Post-Traumatic Stress Disorder (PTSD) \cite{Schwartz2017Aug, Schwartz2018May}, and similar disasters have been associated with subsequent increases in suicides \cite{Krug1998Feb}.

@@

\subsection*{Socio-Econoimc Data}
We used data from the American Community Survey (ACS) administered by the US Census to estimate income levels and the racial composition of neighborhoods where tweets were located.  Data was at the level of the census block group, the smallest unit for which the US Census releases public data.

@@

\section*{Author Contributions}
Author contributions: M.W.C., J.O.-G., A.S. and P.B. designed the research and modeling strategy; A.S. provided data; M.W.C and Z.L. prepared data; M.W.C. and J.O.-G. analyzed data; and M.W.C., J.O.-G., J.L., and P.A.W. wrote the paper.

\noindent This project was co-lead by M.W.C. and J.O.-G. and both contributed equally to this project. Authorship order for these two is listed alphabetically. Authorship order for others listed by proportion of contribution.

>>>>

\begin{abstract}
%Limit 150 words: https://www.nature.com/nclimate/about/content
%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health.  Indeed, traditional analyses of heat and mental health outcomes using data collected by public agencies have found strong associations between elevated temperatures and outcomes such as suicides and mental-health related hospitalizations.  However, these studies, which typically use data reported at the city or county level, have found no difference in vulnerability based on income or race.  We overturn this finding and show that there are stark differences in vulnerability using a novel data source: expressed sentiment in a quarter of a billion geolocated tweets, matched with prevailing weather as well as neighborhood economic and demographic conditions.  We find that increased temperatures worsen expressed sentiment in all areas, but that this effect is much stronger in poor and black neighborhoods.
\end{abstract}
\begin{document}

@@

\section*{Main}
%Overview
Climate change is impacting many aspects of human well-being. A large body of research has explored the ongoing impacts of climate change on agricultural outcomes, economic growth, and physical health \cite{pachauri2014climate}.  However, academic and medical researchers have highlighted the relative paucity of empirical studies on the impacts of climate change on mental health \cite{Berry2018Apr, hayes_climate_2018}. While mental health is under-studied, it accounts for up 13\% of the global burden of disease, representing a major share of total human suffering \cite{Collins2011Jul}, leading to calls for more research into potential linkages between climate change and mental health \cite{Berry2018Apr, Collins2011Jul}. Researchers have begun to answer these calls, finding associations between climate disasters and post-traumatic stress \cite{Waite2017Dec, Raker2019Dec}, deteriorating environmental conditions and a sense of "ecological grief" \cite{Cunsolo2018Apr}, and linkages between higher temperatures and a variety of mental health impacts \cite{baylis_weather_2018, Mullins2019Dec, Li2020Mar, Obradovich2018Oct}.  Nevertheless, with average global temperatures projected to increase by 1.5\textdegree C within a decade \cite{allen2019technical}, more research is needed into the impacts of heat on mental health, especially on identifying the most vulnerable communities.

%Theoretical: frameworks and pathways
Researchers have proposed a number of frameworks that focus on pathways by which temperature can affect mental health \cite{Berry2018Apr, Palinkas2020Apr, BerryETAL2010}. Work on biological mechanisms emphasizes the mental health effects from dehydration and other consequences of maintaining stable body temperature in high heat \cite{Lohmus2018Jul, sadiq_impact_2019}. Additionally, recent studies find nighttime temperatures affect sleep quality, with consequences for mental health \cite{Obradovich2017May, Mullins2019Dec}. Other studies point to linkages between increased temperatures, overall physical health and increases in injuries which can have compounding consequences for exacerbate mental health issues \cite{Berry2007, WHO2007}. Finally, exposure to higher temperatures can affect productivity and income \cite{kjellstrom_impact_2016, Burke2015Nov}, leading to second-order impacts on mental health impacts and outcomes \cite{Katz1997, CohnETAL2004, BouchamaETAL2007}.

%Empirical: Heat and mental health
Results from an emerging literature indicate higher temperatures are also associated with indicators of deteriorating mental health.  Multiple studies find strong evidence that higher temperatures are associated with increases in suicides in the United States \cite{Burke2018Aug, Mullins2019Dec, Dixon2007May}. Other studies find similar relationships in locations around the world \cite{Qi2014Dec, Page2007Aug, Likhvar2011Jan}. Higher temperatures are also correlated with increased hospitalizations related to general mental health issues \cite{Obradovich2018Oct, Mullins2019Dec} and incidents related specific conditions like bipolar disorder and schizophrenia \cite{Lee2007Jan, Sung2013Feb}. Higher temperatures are also linked to increased mortality among people with mental health disorders \cite{Hansen2008Oct}.

The effects of heat exposure on mental health are relatively well established, yet results from research conducted at national scales indicate surprisingly little heterogeneity in impacts, with consequences for our understanding of who may be most or least vulnerable to heat stress and ways we plan for and adapt to changing environmental conditions. For example, a large study of temperature and suicide in the USA and Mexico found ``no significant difference in suicide response to temperature between rich and poor municipalities or counties" \cite{Burke2018Aug}. Similarly, a national study found no effect of income as modifier of the effect of temperature on suicides, emergency department visits, or self-reported mental health status across the USA \cite{Mullins2019Dec}. A major challenge for these studies, however, is their reliance on data from public health agencies, data which is typically aggregated to the municipality or county level.  As a consequence, these studies are often restricted to between-county metrics of vulnerability and cannot take into account neighborhood effects.

Findings from large-scale studies of relatively uniform vulnerability to heat are surprising given minority groups and poorer people are more likely to be exposed to undesirable temperatures and otherwise less able to mitigate the effects of higher heat. For example, people from poorer socio-economic backgrounds are more likely to work outside rather than indoors \cite{Gubernot2014Oct}, more likely to rely on public transportation, bicycling, or walking instead of commuting in their own air-conditioned car \cite{Karner2015Dec}, and more likely to live in housing that is less insulated with poorer quality or no air-conditioning \cite{Samuelson2020Jun}. Thus, there are strong reasons to expect heterogeneities in the impact of heat on mental health outcomes, with these heterogeneities in vulnerability being more pronounced within cities rather than between cities. Indeed, other studies find large differences in vulnerability between neighborhoods for a variety of impacts of heat on physical health  \cite{Belanger2015Mar, Uejio2011Mar} and mental health for other types of climate shocks like hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}.

While public health data on mental health outcomes is not available at the neighborhood scale, spontaneous, in situ data from the social media platform Twitter can provide an important, inexpensive and timely indicator of mental health at extremely precise spatial and temporal scales. Many studies use the mood expressed in social media as an indicator of mental health \cite{Edo-Osagie2020Jul, Sinnenberg2016Dec}. For example, studies have used Twitter data to identify the onset of Post Traumatic Stress Disorder (PTSD) in individuals even before their formal diagnosis \cite{Reece2017Oct}. Similar methods applied to posts on Facebook can predict the onset of depression \cite{Eichstaedt2018Oct}. In London, day-to-day changes in mental health indicators derived from the text of Tweets were associated with changes in mental health crisis episodes \cite{Kolliakou2020Feb}. Across the USA, more positive Tweets are associated with a variety of metrics of human well-being \cite{Mitchell2013May}. Finally, the mood expressed in Tweets has been strongly associated with local weather conditions \cite{baylis_weather_2018, hannak_tweetin_2012} and been used as evidence to support the temperature-suicide relationship \cite{Burke2018Aug}. Thus, posts on Twitter provide an indicator of mental health at fine enough spatial scales to examine neighborhood effects on vulnerability to higher temperatures. We therefore draw on a data set of 243 million geolocated tweets from across the continental United States and covering the period 2009-2019, each paired local weather and neighborhood conditions, to explore how neighborhood conditions affect the relationship between heat and mental heath.

@@

\subsection*{Overall Effect}
We examine the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), a temperature metric that accounts for dry-bulb temperature, humidity, wind speed, and solar radiation to more accurately describe the effects of heat on the human body \cite{budd2008wet}. Controlling for a variety of fixed effects across all Tweet, we find higher temperatures are associated with lower sentiment (i.e., worsening mood; See Fig. \ref{fig:wbgt}).  Sentiment is highest at 5\textdegree C WBGT (a dry bulb temperature typically around 12\textdegree C/54\textdegree F) with the largest declines in sentiment between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F). Sentiment also declines with colder temperatures (below 5\textdegree C WBGT), although only slightly. We ran models using multiple measures of sentiment and find similar results and patterns across all sentiment measures (see Supplement).

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{../res/wbgt.png}
  \caption{Relationship between Wet Bulb Globe Temperature (WBGT) and sentiment from text in Tweets.  As temperatures increase above 5\textdegree C WBGT, sentiment rapidly declines.}
  \label{fig:wbgt}
\end{figure}

We find median neighborhood income strongly moderates the relationship between temperature and sentiment (see Fig \ref{fig:hetero}, panel A), with large differences in sentiment between the poorest and wealthiest neighborhoods. As temperatures increase to a modest 20\textdegree C WBGT (29\textdegree C/84\textdegree F), sentiment increases in the wealthiest neighborhoods (95th income percentile) but decreases in the median and lowest income neighborhoods. The wealthiest neighborhoods do not see decreases in sentiment until temperatures exceed 20\textdegree C WBGT (29\textdegree C/84\textdegree F), at which point sentiment decreases relatively evenly regardless of neighborhood income percentile.

We explored the relationship between neighborhood racial group characteristics and temperature and sentiment and find the effects of heat are felt disproportionately in neighborhoods that are majority black (see Fig \ref{fig:hetero}). Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the sentiment of tweets in majority black neighborhoods decrease four times as much as the sentiment of people in other neighborhoods. Additionally, at modest to moderate temperatures of 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority Hispanic neighborhoods have slightly lower sentiment than people in majority white or other neighborhoods, although this gap narrows at higher temperatures.

Exploring the combined effects of neighborhood percent minority race and income in moderating the impacts of heat on mental health, we find separate effects for these two variables (see Supplement). In other words, neither neither percent minority population nor income alone account for the heterogeneity in vulnerability as neighborhoods that are both poorer and higher percent black are more affected by heat than than neighborhoods that are just poorer or higher proportion black.

@@

\subsection*{Sentiment Comparison With Other Events}
Expressed sentiment is a widely used metric to assess mental health and well-being on social media.  While a variety of related metrics can be used to quantify sentiment, we use the VADER metric that was specifically designed for microblogs like Twitter \cite{hutto2014vader}.  To give context to this novel metric, we compare the impacts of heat waves, defined as a change from 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), to two other changes that have large impacts on sentiment (See Fig. \ref{fig:compare}).  First, we examine the change in average sentiment from Saturday to Monday, the highest and lowest days for Twitter sentiment as well as other mental health metrics, such as suicide rates \cite{CDC2021}.  We also examine the decrease in sentiment associated with one of the most expensive hurricanes in the last decade in the United States: Hurricane Sandy.  Specifically, we compare the mean sentiment of counties affected by Sandy on the week after landfall to the mean sentiment for the week previous.  Hurricanes like Sandy are associated with mental health effects including Post-Traumatic Stress Disorder (PTSD) \cite{Schwartz2017Aug, Schwartz2018May}, and similar disasters have been associated with subsequent increases in suicides \cite{Krug1998Feb}.

@@

\subsection*{Socio-Econoimc Data}
We used data from the American Community Survey (ACS) administered by the US Census to estimate income levels and the racial composition of neighborhoods where tweets were located.  Data was at the level of the census block group, the smallest unit for which the US Census releases public data. Following Census categories, we report racial characteristics of neighborhoods as majority population of four broad racial categories - non-Hispanic black, non-Hispanic white, Hispanic of any race, and other, which includes Native American, multi-racial, Asian-American and Pacific Islander.

@@

\section*{Author Contributions}
Author contributions: M.W.C., J.O.-G., A.S. and P.B. designed the research and modeling strategy; A.S. provided data; M.W.C and Z.L. prepared data; M.W.C. analyzed data; and M.W.C., J.O.-G., J.L., and P.A.W. wrote the paper.

\noindent This paper was the result of a SESYNC graduate pursuits project co-lead by M.W.C. and J.O.-G.
